On Optimality of Finite Element Integration

### 2016-01-20 TOMS_Paper/assembly.tex
Went on adding stuff to formalization

\item If $\mathbb{S} = \mathbb{T}$, jump to step (6). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^1, ..., T^n \rbrace$.\\ \textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}. 
\item Only one terminal $T$ adjacent to $S$: apply strategy (ii) including one non-terminal symbol $S^{nt}$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^{nt} \rbrace$.
\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii).
\end{enumerate}
Go back to step (2).

@@

\section{Formalization}
\label{sec:proof}
We demonstrate that the orchestration of sharing elimination and pre-evaluation performed by the transformation algorithm guarantees quasi-optimality (Definition~\ref{def:mln-quasi-optimality}). The proof re-uses concepts and explanations provided throughout the paper, as well as the terminology introduced in Section~\ref{sec:se-algo}.

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$. If $T_{m_{1}} \cap\ T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary. $\Gamma$, therefore, may skip a potential global optimum, since $\theta$ only reasons on monomials in ``isolation''. We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a local -- instead of a global -- optimum. This situation is prevented by construction, because $\Gamma$ analyzes all possible bipartitions searching for a global optimum preserving Constraint~\ref{const:TH}. Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished. Finally, note how this problem can be seen as an instance of the well-known Knapsack problem.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of the normal form of expressions and the structure inherent in tensor operations involving test and trial functions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (strategy (i)) or factorization of multilinear symbols (achieved through strategy (ii)). If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration. Since strategy (i) is applied if and only if $|P| < |S_P|$ and requires neither expansion nor factorization, step (3) will not prune any optimal solution from the search space; in fact, it simply represents a step forward towards quasi-optimality.

Step (6) factorizes symbols following the ordering dictated by the sharing graph and its rewrite rules. Non f-decreasing factorizations are excluded by construction (step (5)). We now have to prove that applying in sequence rewrite rules (A), (B), and (C), with applications of rule (C) that may expose further opportunities for (A) and (B), always represents progress towards quasi-optimality. 
\begin{description}
\item[Rule (A)] $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule is side effects free (i.e., no need to choose between different factorizations).
\item[Rule (B)] Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$. Let us consider a generic alternative factorization $S^{nt}(...S + ... + ...V)$. If $V$ were a terminal, we would be back to where we started, because rule (B) with $V = T$ and $S = S^{nt}$ would still be applicable. Otherwise ($V$ non terminal), we would have $S$ appearing in two different sub-expressions, which would imply sub-optimality as a consequence of spatial sharing.
\item[Rule (C)] We can interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively. For each disconnected component $\mathbb{B}^d = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$ in $\mathbb{G}$, we consider two possibilities, namely $\mathbb{B^d}$ 1) is not or 2) is complete. Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), the lower bound on the operation count (as number of products performed at every loop iteration), $lb$, is given by $lb = {min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$. This is reachable by factorizing the vertex with highest degree, as any other choice may break factorization opportunities. In case 2), we have $lb = 1$ because factorization is applicable iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$); this is handled by going back to step (2), at the end of step (6).
\end{description}
The algorithm is guaranteed to terminate because of the jump to step (6) when $\mathbb{S} = \mathbb{T}$. 
\end{enumerate}
\end{proof}

%TODO : In addition, if X holds, then $\Lambda'$ is optimal as established by Definition ?.
%TODO : A list of preconditions that eventually result in optimality

>>>>

\item If $\mathbb{S} = \mathbb{T}$, jump to step (6). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii).\\
\textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}.
\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply strategy (ii).\\
\textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}. 
\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii) such that all symbols $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$ are grouped.\\
\textit{Note: this recasts the expression as $S (...S^1 + ... + ...S^n)$}.
\end{enumerate}
Go back to step (2).

@@

\section{Formalization}
\label{sec:proof}
We demonstrate that the orchestration of sharing elimination and pre-evaluation as performed by the transformation algorithm guarantees quasi-optimality (Definition~\ref{def:mln-quasi-optimality}). The proof re-uses concepts and explanations provided throughout the paper, as well as the terminology introduced in Section~\ref{sec:se-algo}.

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$. If $T_{m_{1}} \cap\ T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary. $\Gamma$, therefore, may skip a potential global optimum, since $\theta$ only reasons on monomials in ``isolation''. We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a local -- instead of a global -- optimum. This situation is prevented by construction, because $\Gamma$ analyzes all possible bipartitions searching for a global optimum preserving Constraint~\ref{const:TH}. Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished. Finally, note how this problem can be seen as an instance of the well-known Knapsack problem.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of the normal form of expressions and the structure inherent in tensor operations involving test and trial functions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (strategy (i)) or factorization of multilinear symbols (achieved through strategy (ii)). If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration. Since strategy (i) is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) will not prune optimal solutions from the search space.

Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules. Non f-decreasing factorizations are excluded by construction at step (5). We then prove that applying in sequence rewrite rules (A), (B), and (C), with applications of rule (C) that may expose further opportunities for (A) and (B), represents a step forward towards quasi-optimality. 
\begin{description}
\item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
\item[Rule (B)] Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$. Let us consider the alternative factorization $S^{nt}(...S + ... + ...V)$. If $V$ were a terminal, it is easy to see that setting $V = T$ and $S = S^{nt}$ would just expose another possible hit for rule (B). Otherwise ($V$ non terminal), we would have $S$ now appearing in two different sub-expressions ($(...S + ... + ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$. From this, we see that applying rule (C) prior to rule (B) may prune the optimal solution from the search space.
\item[Rule (C)] We interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively. For each disconnected component $\mathbb{B}^d = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$ in $\mathbb{G}$, we consider two possibilities, namely $\mathbb{B^d}$ 1) is not or 2) is complete. Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), and thanks to loop linearity, the lower bound on the operation count (as number of products performed at every loop iteration), $lb$, is given by $lb = {min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$. This is reachable by factorizing the vertex with highest degree, as any other choice may break factorization opportunities. In case 2), we have $lb = 1$ because factorization is applicable iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$). This lower bound is achieved by repeating the sharing elimination algorithm (notice the jump back to step (2) at the end of step (6)).
\end{description}
The algorithm is guaranteed to terminate because of the jump to step (6) as $\mathbb{S} = \mathbb{T}$. 
\end{enumerate}
\end{proof}

%\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii)
%\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply strategy (ii).\\
%\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii) such that all symbols $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$ are grouped.\\

Throughout the paper we have reiterated the idea that Algorithm~\ref{algo:gamma} reaches optimality if some preconditions are satisfied by the input variational form. We here discuss such preconditions, in increasing order of complexity.
\begin{enumerate}
\item There is a single monomial and only a specific coefficient, the coordinates field. This is by far the simplest scenario, which requires no particular transformation at the level of the outer loops, so optimality naturally follows.
\item There is a single monomial, but multiple coefficients are present. Optimality is achieved if only structured tensor operations are present (see Section~\ref{sec:se-rln}), since there is no ambiguity in factorization, which is exploited during step (7) of Algorithm~\ref{algo:sharing-elimination}.
\item There are multiple monomials, but no coefficients other than the coordinates field are present, or there is no spatial sharing in coefficients across different monomials. This reduces to cases (1) and (2) above.
\item There are multiple monomials and there is sharing of coefficients across different monomials. We can still in some cases ....
\end{enumerate}
